Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-10-17 sports_analytics.Rmd
added `nflplotR` figure (#31)

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
library(tidyverse)
library(kableExtra)
```

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR",
    "sportyR"
  ),
  file = "pkgs.bib"
)

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")

@@

To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
Since both ratings are below 2400, we apply a development coefficient of $K=20$ for this example.
The probability of author ($a$) defeating their opponent ($b$) is
<!-- don't know how to include `r` inside latex -->
$$
\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
$$
The author that played this game ended up with a victory, which means the associated score is $S_a = 1$.
The post-game Elo rating for this author is
$$
R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
$$

@@

>>>>

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
library(tidyverse)
theme_set(theme_bw())
library(kableExtra)
```

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR",
    "sportyR",
    "nflplotR"
  ),
  file = "pkgs.bib"
)

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme(legend.position = "top")

@@

To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
Since both ratings are below 2400, we apply a development coefficient of $K=20$ to this example.
The probability of author ($a$) defeating their opponent ($b$) is
<!-- don't know how to include `r` inside latex -->
$$
\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
$$
The playing author finished with a victory, which is associated with a score of $S_a = 1$.
The post-game Elo rating for this author is
$$
R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
$$

@@

Sharing a similar goal with \pkg{teamcolors} is \pkg{nflplotR} [@R-nflplotR], which also provides \pkg{ggplot2} extensions but is designed specifically for the NFL.
A great feature of \pkg{nflplotR} is the \pkg{ggplot2} "geom" (geometric object) functions that enhance high-quality plotting of NFL team logos and player images.
Figure \ref{fig:nflplotr} shows a scatterplot of offensive and defensive expected points added for NFL teams in the 2021 regular season.
The logos of all 32 American football clubs are plotted in place of the usual dots, giving a better description of which team each data point represents.

```{r nflplotr, fig.cap="Offensive and defensive per play for the 2021 NFL regular season, plotted with \\pkg{nflplotR} using data from \\pkg{nflfastR}."}
pbp <- nflreadr::load_pbp(2021) %>% 
  dplyr::filter(season_type == "REG") %>%
  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))

offense <- pbp %>%
  dplyr::group_by(team = posteam) %>%
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))

defense <- pbp %>%
  dplyr::group_by(team = defteam) %>%
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))

combined <- offense %>%
  dplyr::inner_join(defense, by = "team")

combined |> 
  ggplot(aes(off_epa, def_epa)) +
  nflplotR::geom_mean_lines(aes(v_var = off_epa , h_var = def_epa)) +
  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.08, alpha = 0.8) +
  labs(x = "Offensive EPA/play",
       y = "Defensive EPA/play") +
  scale_y_reverse()
```
